---
title: "HW 2 Data Science"
author: "Victoria Mello (vsm2118)"
date: "October 4, 2023"
output: github_document
---
# Promblem 1

```{r}
library(tidyverse)
```

Clean and tidy the datasets: 

```{r clean_538_pols}
month_df = 
  tibble(
    month_num = 1:12,
    month_abb = month.abb,
    month = month.name)

pols = 
  read_csv("./data/pols-month.csv") |>
  separate(mon, into = c("year", "month_num", "day"), convert = TRUE) |>
  mutate(
    president = recode(prez_gop, "0" = "dem", "1" = "gop", "2" = "gop")) |>
  left_join(x = _, y = month_df) |> 
  select(year, month, everything(), -day, -starts_with("prez")) 
```

```{r clean_538_snp}
snp = 
  read_csv("./data/snp.csv") |>
  separate(date, into = c("month", "day", "year"), convert = TRUE) |>
  arrange(year, month) |>
  mutate(month = month.name[month]) |>
  select(year, month, close) 
```

```{r clean_538_unemp}
unemployment = 
  read_csv("./data/unemployment.csv") |>
  rename(year = Year) |>
  pivot_longer(
    Jan:Dec, 
    names_to = "month_abb",
    values_to = "unemployment"
  ) |> 
  left_join(x = _, y = month_df) |> 
  select(year, month, unemployment)
```

Now we merge the three datasets!

```{r merge_538}
data_538 = 
  left_join(pols, snp) |>
  left_join(x = _, y = unemployment)

str(data_538)
```

The `pols` data has `r nrow(pols)` observations and `r ncol(pols)` variables and tells us about the party affiliation distribution (democrat or republican) for governors and senators for a given year from years `r range(pols$year)[1]` to `r range(pols$year)[2]`. It also tells us whether the sitting president was a democrat or republican. The `snp` data has `r nrow(snp)` observations and `r ncol(snp)` variables, ranging from years `r range(snp$year)[1]` to `r range(snp$year)[2]`. The `unemployment` data has `r nrow(unemployment)` observations and `r ncol(unemployment)` variables ranging from years `r range(unemployment$year)[1]` to `r range(unemployment$year)[2]`. In Januarys in or after 1975 in which a democrat was president, the **average unemployment rate was `r filter(data_538, month == "January", year >= 1975, president == "dem") |> pull(unemployment) |> mean() |> round(2)`**.  The average unemployment rate over the same time period in which a republican was president was `r filter(data_538, month == "January", year >= 1975, president == "gop") |> pull(unemployment) |> mean() |> round(2)`.

# Problem 2

```{r}

# Step 1: Read and clean the Mr. Trash Wheel dataset
mr_trash_wheel <- readxl::read_excel("./trashwheel.xlsx", sheet = "Mr. Trash Wheel") %>%
  select(Dumpster, Month, Year, Date, `Weight (tons)`, `Cigarette Butts`, `Homes Powered*`) %>%
  rename(
    Weight_tons = `Weight (tons)`,
    Cigarette_Butts = `Cigarette Butts`,
  ) %>%
  filter(!is.na(Dumpster)) %>%
  mutate(Month = tolower(Month))  

mr_trash_wheel <- mr_trash_wheel %>%
  mutate(
    Year = as.integer(Year),
    homes_powered = (Weight_tons * 500) / 1000)

# Step 2: Import, clean, and organize data for Professor Trash Wheel
professor_trash_wheel <- readxl::read_excel("./trashwheel.xlsx", sheet = "Professor Trash Wheel") %>%
  select(Dumpster, Month, Year, Date, `Weight (tons)`, `Cigarette Butts`) %>%
  rename(
    Weight_tons = `Weight (tons)`,
    Cigarette_Butts = `Cigarette Butts`
  ) %>%
  filter(!is.na(Dumpster)) %>%
  mutate(Month = tolower(Month), Trash_Wheel = "Professor Trash Wheel")

professor_trash_wheel <- professor_trash_wheel %>%
  mutate(
    Year = as.integer(Year),
    homes_powered = (Weight_tons * 500) / 1000)

# Step 3: Import, clean, and organize data for Gwynnda
gwynnda <- readxl::read_excel("./trashwheel.xlsx", sheet = "Gwynnda Trash Wheel") %>%
  select(Dumpster, Month, Year, Date, `Weight (tons)`, `Cigarette Butts`, `Homes Powered*`) %>%
  rename(
    Weight_tons = `Weight (tons)`,
    Cigarette_Butts = `Cigarette Butts`,
    Homes_Powered = `Homes Powered*`
  ) %>%
  filter(!is.na(Dumpster)) %>%
  mutate(Month = tolower(Month), Trash_Wheel = "Gwynnda")

gwynnda <- gwynnda %>%
  mutate(
    Year = as.integer(Year),
    homes_powered = (Weight_tons * 500) / 1000)

```

```{r}
# Step 4: Combine the datasets
combined_data <- bind_rows(mr_trash_wheel, professor_trash_wheel, gwynnda)
view(combined_data)
```

The resulting dataset has `r nrow(combined_data)` observations, which include data from Mr. Trash Wheel, Professor Trash Wheel, and Gwynnda. Some key variables in the dataset are Dumpster, Month, Year, Date, Weight_tons, Cigarette_Butts, and homes_powered. The total weight of trash collected by Professor Trash Wheel is `r sum(professor_trash_wheel$Weight_tons)` tons. In July of 2021, Gwynnda collected  `r sum(gwynnda$Cigarette_Butts[gwynnda$Month == "july" & gwynnda$Year == 2021])` cigarette butts.



